---
title: "WORKSHOP 1: FOUNDATIONS OF DATA SCIENCE"
author: "Giuliana Ponce"
format:
  html: 
    code-fold: true
    code-summary: "Show code"
    code-tools: true
editor: visual
---
::: {.callout-note collapse="true" appearance="minimal" icon="false"}
## NOTES

### 1. Ingesting data into R

Loading the primary data science framework and Excel import library

```{r}

library(tidyverse)
library (readxl)
```

**Practice Import A: Loading a standard coma-separated plain text file**

We have to make sure that we tell R exactly where our file is because if not it's going to show us an error (ours is in the data folder, inside of the workshop1 folder; if we don't put exactly that, R cannot find the file).

```{r}
benthic_cover <- read_csv(here::here("data/workshop1/reef_cover_log.csv"))
```

**Practice Import B: Parsing a tab-separated telemetry instrument array string**

```{r}
acoustic_stream <- read_tsv(here::here("data/workshop1/acoustic_telemetry_stream.txt"))
```

**Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet**

We add at the end the "sheet" section with the name of the spreadsheet that we want

```{r}
fisheries_annual <- read_excel(here::here("data/workshop1/fish_catch_data.xlsx"), sheet ="Commercial_2026")
```

When we work with excel we tend to make notes and R treats this conversational titles as headers corrupting the entire data

```{r}
mangrove_data <- read_csv(file=here::here("data/workshop1/mangrove_survey_raw.csv"))
```

We can use arguments to skip headers and we can declare missing flags

```{r}
mangrove_data <- read_csv(
  here::here("data/workshop1/mangrove_survey_raw.csv"),
  skip = 5,   # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA
```

### 2. Data frame architectures

Force a modern tibble to degrade into a legacy base R data frame structure

```{r}
benthic_cover_df <- as.data.frame(benthic_cover)
```

and we compare

Print the old-style dataframe structure to view

```{r}
print(benthic_cover_df)
```

And compare with tibble alternative

```{r}
print(benthic_cover)
```

### 3. **Wrangling out ecological signals using Palmer Penguins**

To explore how data wrangling reveals information about underlying ecology, we will transition to working with a landmark built-in dataset: the Palmer Penguins archive.

First of all we need to install and load the package ("palmerpenguins").

```{r}
library("palmerpenguins")
data("penguins")
```

We need to examine the structure of the dataset (ALWAYS DO THIS!!!!)

```{r}
glimpse(penguins)
str(penguins)
```

The glimpse command maps out the entire architectural anatomy of our dataset: <fct> (Factor): Categorical groupings with fixed levels (e.g., species containing Adelie, Chinstrap, and Gentoo). <dbl> (Double): Continuous numeric measurements containing decimals (e.g., bill_length_mm). <int> (Integer): Whole number variables tracking counts (e.g., body_mass_g).

We also need to generate the exploratory summary matrix

```{r}
summary(penguins)
```

This is an immediate diagnostic tool That reports the presence of missing indicators (NA).

### 4. Foundational grammar

#### 4.1 Isolating attributes with select()

Ecological data tables frequently collect duplicate logistics records that do not contribute to morphological or physiological questions. The select() verb allows you to slice your datasets vertically, isolating or dropping columns based on variable names:

-   Vertically slice specific morphometric variables by explicit name:

```{r}
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)
```

-   Retain a continuous block of attributes using the colon operator

```{r}
spatial_block <- select(penguins, species:island)
glimpse(spatial_block)
```

-   Discard logistics tracking attributes while preserving everything else using the minus sign

```{r}
clean_scientific_fields <- select(penguins, -year)
glimpse(clean_scientific_fields)
```

#### 4.2 Sifting rows with select()

While select() targets variables vertically, filter() isolates records horizontally based on targeted conditional parameters. R evaluates every row against your logical expressions, retaining entries that evaluate to TRUE and dropping those that return FALSE or NA

-   Isolate observations belonging to a single categorical target group:

```{r}
adelie_cohort <- filter(penguins, species == "Adelie")
```

-   Sift out individuals using continuous numerical boundary thresholds.

```{r}
heavy_penguins <- filter(penguins, body_mass_g > 4500)
#Preserves only large penguins whose mass exceeds 4500 grams
```

-   Combine multiple conditional parameters across separate attributes

```{r}
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")
# Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
```

-   Sift records matching multiple targeting flags within an explicit set

```{r}
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

#### 4.3 Ordering sequences with arrange()

The arrange() verb alters the sorting configuration of rows within the data frame without changing individual cell values. This is essential for inspecting size hierarchies or chronologically ordering long-term environmental observation lines:

-   Sort penguins by ascending body mass (Default setting: Smallest mass first)

```{r}
lightest_first <- arrange(penguins, body_mass_g)
```

-   Sort penguins in descending sequence using the desc() layout wrapper

```{r}
heaviest_first <- arrange(penguins, desc(body_mass_g))
```

-   Execute nested sorting criteria: Group by species, then sort by descending bill length

```{r}
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

#### 4.4. Pipe (\|\>)

Pipe helps us to pass our data through a series of transformations in a single linear flow, avoiding creating multiple versions of our data.

**The Logic of the Pipe:**

-   **Without a Pipe:** You have to read the code "inside-out." You start in the middle, perform the mutate(), then wrap that in a filter(), and finally wrap that all in a select().

-   **With a Pipe:** You read from left to right, or top to bottom. You take your data, then you filter it, then you mutate it to create a new column, then you select the columns you need.

*Example:*

-   Without a Pipe:

```{r}
penguins_subset <- mutate(penguins, bill_ratio = bill_length_mm / bill_depth_mm)
penguins_final <- filter(penguins_subset, species == "Adelie")
```

-   With a Pipe:

```{r}
penguins_final_PIPE <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```

By using the pipe, our code becomes easier to read, easier to debug, and much more efficient to write!

#### 4.5. **Computing new attributes with mutate()**

We use the mutate() verb to modify existing attributes or append entirely new vectors to the data frame.

-   For example, for calculating a new morphological ratio in our environment:

```{r}
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )

glimpse(penguin_ratios)
```

### 5. **Data aggregation and ecological summarisation**

To extract meaningful biological conclusions (such as determining whether morphological allometry varies by species or island group) you must compress thousands of individual observations into explicit population summary metrics.

This is achieved using the dual-layered combination of group_by() and summarise().

When you apply group_by(), R does not alter the physical appearance of the table. Instead, it alters how it interacts with subsequent functions by creating hidden, virtual data buckets based on categorical factor levels. When you immediately follow this with summarise(), R calculates targeted statistical reductions for each individual bucket and collapses them into a brand-new summary tibble.

*Example:*

-   Grouping our active memory penguins by species

```{r}
grouped_penguins <- group_by(penguins, species)
#Notice that the table looks identical, but metadata notes 'Groups: species [3]'
print(grouped_penguins)
```

-   Collapsing the buckets into explicit summary metrics

```{r}
species_mass_summary <- summarise(grouped_penguins,
  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
```

If a column contains even a single missing observation flag (NA), any mathematical aggregation function (such as mean(), sd(), or sum()) will automatically return NA. To override this safely, you must explicitly declare that missing cells should be dropped during calculation.

-   Overcoming the missing value trap using na.rm = TRUE

```{r}
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)
```
:::
## CHALLENGES:
